---
title: "Behavioural Principles Behind SRTM"
author: "SRTM package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Behavioural Principles Behind SRTM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The **SRTM** (Self-Referenced Trajectory Modelling) framework was built to
answer a practical question:

> *How can we evaluate the impact of educational influences in real classrooms,
> without relying on idealised randomised controlled trials (RCTs) that are
> rarely feasible in practice?*

Instead of comparing a treatment group to a control group, SRTM treats
**each student as their own control**, and uses their personal trajectory over
time as the reference point for evaluating change.

This vignette focuses on the **behavioural principles** underpinning SRTM and
how they translate into statistical and modelling decisions in the package.

We assume you have already seen the basic workflow in the
`Getting Started with SRTM` vignette.

# 1. Why not RCTs in classrooms?

RCTs are powerful in tightly controlled environments (e.g., some areas of
medicine), but classroom contexts pose several challenges:

- **Random assignment** is often infeasible or unethical:
  you cannot easily randomise students to "good teaching" vs "no teaching".
- There is rarely a clear **standard treatment** to act as a control
  condition.
- The **sample sizes** and number of confounding variables
  (between-student, between-classroom, between-school) make it difficult
  to obtain clean contrasts.
- Educational systems are **dynamic and context-sensitive**: rigid control
  structures can erode authenticity and teacher autonomy.

SRTM is a pragmatic response to these realities. Instead of forcing classroom
data into an RCT mould, it builds on a set of behavioural principles that
respect complexity while still permitting systematic analysis.

# 2. Kennedy's Behavioural Principles

SRTM is guided by a set of behavioural principles which, if not found
elsewhere in the literature, you may treat as **Kennedy's Behavioural
Principles**.

They are purposely informal: they are not offered as universal laws of human
behaviour, but as useful working assumptions for educational evaluation over
short-to-medium time scales.

## 2.1 0th Law – Behavioural Complexity

> *Any observed behaviour is the result of a complex, dynamic network of
> attitudes, values, beliefs, and contextual factors.*

Key points:

- There is a large, effectively uncountable number of "weights" connecting
  internal states (attitudes, values, beliefs) to behaviour.
- These weights are **not constant**:
  they depend on the current state of the person, the context, and time.
- They are **recursive and self-updating**: internal states influence how new
  information is perceived, which in turn updates those same states.

From a modelling point of view:

- attempting to estimate these weights directly is not tractable;
- SRTM therefore works with **observable measures** (e.g., attitude scores,
  engagement ratings, achievement) and their **trajectories over time**.

The 0th Law encourages us to be modest:
we accept that our variables are **proxies** for a much richer process.

## 2.2 1st Law – Behavioural Inertia

> *In the absence of a net external influence, the rate at which attitudes,
> values and beliefs change is essentially constant over short periods in
> relatively stable contexts.*

"Essentially" is doing a lot of work here:

- internally, a great deal may be happening;
- but at the level of our measurement tools, the *trend* in a given variable
  (e.g., an attitude score) can be approximated as a straight line over a
  short enough window.

In SRTM this motivates the idea of a **personal trajectory**:

- use two time points, `y0` (historical) and `y1` (baseline), to estimate each
  student's slope of change up to the onset of an influence;
- assume that, if nothing new happened, this slope would have continued in a
  similar way to `y2`.

This is what allows us to model the **counterfactual** – what would likely
have happened for *this student* if the new influence had not occurred.

## 2.3 2nd Law – Behavioural Momentum

> *When a net external influence is present, the rate of change in behaviour
> is approximately proportional to the perceived strength of the influence and
> inversely proportional to behavioural inertia.*

Implications:

- Students with **well-established** attitudes and values (high inertia) are
  harder to shift.
- Students whose internal systems are **less settled** change more readily.
- What matters is **perceived** influence:
  the messages received, not just the messages sent.

SRTM uses this principle to interpret systematic deviations from the
counterfactual:

- if a group of students is consistently **above** its expected trajectory,
  this is consistent with the presence of a meaningful positive influence;
- if a group is consistently **below** expectation, this suggests that
  something is inhibiting progress.

The quantitative tools (regression, grouping, t-tests) are conventional; the
interpretation is framed by behavioural momentum.

## 2.4 3rd Principle – Behavioural Impulse

> *Influences rarely act as one-off events; instead, they operate over time.
> The overall effect depends on both perceived strength and duration.*

Conceptually this is analogous to impulse in physics:

\[
\text{Impulse} \approx \text{Force} \times \text{Time}.
\]

In educational contexts:

- A pedagogical change applied for **too short** a duration may be
  indistinguishable from noise (especially if the measurement tool is coarse).
- Over **very long** durations, other influences (school changes, life events,
  curriculum shifts) accumulate, making attribution more ambiguous.

SRTM does not solve this trade-off, but it makes it explicit:

- the user must choose time gaps (or dates) for `y0`, `y1`, `y2`;
- the method is most defensible when these are short enough that the 1st Law
  (behavioural inertia) is plausible, but long enough for meaningful change
  to be observable.

# 3. From principles to models

How do these principles show up in the SRTM workflow?

## 3.1 Personal trajectories

Given measurements at `y0`, `y1`, `y2` and time gaps `dt01`, `dt12`:

```r
library(SRTM)
data("SRTM_synth_data")

x <- SRTM_synth_data
head(x)
```

SRTM:

1. Uses `y0` and `y1` (and `dt01`) to compute a **slope** for each student:

   ```r
   # behind the scenes in SRTMAnalyse():
   # m01 <- calculateTrajectories(data = x, y0 = "y0", y1 = "y1", time_period = dt01)
   ```

2. Treats this slope as a summary of the student's **behavioural inertia**
   in that measure.

This step is where the 1st Law (behavioural inertia) is operationalised:
a linear trend is used as a local approximation to a more complex underlying
process.

## 3.2 Grouping students

SRTM then groups students in two complementary ways:

- by **trajectory** (slope between `y0` and `y1`), and  
- by **baseline level** (measurement at `y1`).

The grouping is typically done via `findGroups()` and `assignGroups()`:

```r
# Conceptual only; in practice SRTMAnalyse() orchestrates these calls
traj_params <- findGroups(x, time_var = "m01", interactive = FALSE)
traj_group  <- assignGroups(x, group_params = traj_params)

base_params <- findGroups(x, time_var = "y1", interactive = FALSE)
base_group  <- assignGroups(x, group_params = base_params)
```

This yields groups like:

- "low, stable"  
- "low, improving"  
- "high, declining"  
- etc.,

depending on the data and the choices made in `findGroups()`.

These groups represent **types of student trajectories**, not fixed
psychological categories.

## 3.3 Expected vs observed post scores

For each baseline × trajectory group, SRTM then fits simple linear models
to predict what would likely happen at `y2`:

```r
# Conceptual only:
# df <- predictPostResponse(data = x, y0 = "y0", y1 = "y1", y2 = "y2",
#                           base_group = "baseGroup", traj_group = "trajGroup",
#                           time01 = dt01, time12 = dt12)
```

The predicted value is stored as `exp_y2`.

This embodies the question:

> *If nothing new had happened, and the world behaved as it did between
> y0 and y1, where would this student (and group) likely be at y2?*

## 3.4 Comparing with reality

Finally, SRTM compares `y2` (observed) with `exp_y2` (expected):

```r
# part of SRTMAnalyse():
# res <- compareOutcomes(data = df, obs = "y2", exp = "exp_y2",
#                        baseGroups = "baseGroup", trajGroups = "trajGroup")
```

At group level, we look at:

- the mean difference (`mean(y2 - exp_y2)`),
- its standard error and t-statistic,
- and the corresponding p-value.

From the standpoint of the behavioural principles:

- a **large, consistent positive difference** suggests a net positive
  influence strong enough to overcome behavioural inertia;
- a **large negative difference** suggests that something has disrupted or
  reversed the prior trend;
- small differences are consistent with continuation of the pre-existing
  trajectory.

SRTM does not insist that these differences be interpreted causally in a
strong sense; rather, it presents them as structured evidence for
professional judgement.

# 4. Implications for practice

A key design philosophy in SRTM is:

> *Support teachers to interpret change in their own context, rather
> than prescribing a single "best practice".*

Some practical implications:

- Different **groups of students** may respond differently to the same
  influence; SRTM is built to highlight this heterogeneity.
- Negative or null results for some groups are not necessarily failures;
  they can be prompts for local sense-making (e.g., "this worked well
  for already-engaged students, but not for those starting low").
- The method is **transparent**:
  there is no black-box model; the core pieces are linear trends,
  clustering, and t-tests, framed by behavioural principles.

# 5. A minimal worked example

For completeness, here is a compact example that you can run:

```{r}
library(SRTM)

data("SRTM_synth_data")
x <- SRTM_synth_data

# Run analysis non-interactively for the vignette
res <- SRTMAnalyse(
  data        = x,
  y0          = "y0",
  y1          = "y1",
  y2          = "y2",
  id_col      = "ID",
  time01      = 1,
  time12      = 1,
  group_first = "baseline",
  interactive = FALSE
)

# Group-level comparisons
res$Comparisons
```

You can then use `summary(res)` and `plot(res)` in an interactive R session
to explore the same results more visually.

# 6. Conclusion

SRTM is not a replacement for all forms of evaluation, nor is it a magic
detector of "what works". Its aim is more modest and, we hope, more useful:

- to provide a coherent way to think about change over time in real classrooms,
- to treat each student as their own reference point,
- to visualise and test how different *kinds* of students respond, and
- to support professional interpretation rather than override it.

For technical details on implementation, see the documentation for:

- `SRTMAnalyse()`
- `assignGroups()`
- `compareOutcomes()`
- `summariseGroupOutcomes()`
- `SRTM_student_attitude_data`

