---
title: "Getting Started with SRTM"
author: "SRTM package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with SRTM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The **SRTM** (Self-Referenced Trajectory Modelling) package provides tools
for evaluating change in real classrooms by treating **each student as their
own control**.

Rather than relying on randomised control trials (RCTs), SRTM uses three
measurement occasions

- `y0`: historical (pre-pre) or background measure  
- `y1`: baseline or pre-influence measure  
- `y2`: post-influence measure  

to estimate what would likely have happened for each student **if nothing new
had happened**, and then compare this to what actually happened.

This vignette walks through a quick, teacher-friendly workflow using the
included synthetic example data.

# 1. Setup

Make sure the package is installed and loaded.

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("DrJPK/SRTM")
```

```{r, message = FALSE}
library(SRTM)
```

# 2. Data: synthetic example

The package includes a synthetic dataset that mimics real classroom data.

```{r}
data("SRTM_synth_data")
x <- SRTM_synth_data

str(x)
head(x)
```

You should see at least three numeric columns (`y0`, `y1`, `y2`) and an ID
column (e.g. `ID` or `participantID`).

If you have your own Excel file in the appropriate format, you can instead use:

```{r, eval = FALSE}
x <- importSRTMExcel("my_class_data.xlsx")
```

# 3. Running SRTMAnalyse()

The main entry point is `SRTMAnalyse()`. For basic usage with the demo data:

```{r}
res <- SRTMAnalyse(
  data        = x,
  y0          = y0,
  y1          = y1,
  y2          = y2,
  id_col      = "ID",
  group_first = "baseline",
  interactive = FALSE
)
```

In interactive use (e.g. in the R console), you can set `interactive = TRUE`.
This will:

- prompt for time intervals or dates,  
- show plots of distributions,  
- and ask you to confirm grouping decisions.

In vignette mode we set `interactive = FALSE` so the document can knit
non-interactively.

You can inspect the resulting object:

```{r}
res
str(res, max.level = 1)
```

The object `res` has class `"srtm_analysis"` and contains:

- `res$data`: the augmented analysis dataset  
- `res$Comparisons`: observed vs expected comparisons by group  
- `res$GroupSummary`: descriptive summaries by group  
- `res$settings`: analysis settings (time intervals, grouping order, etc.)  
- `res$group_params`: details of the group-finding steps  

# 4. Visualising results

If a plotting method is available, you can call:

```{r, eval = FALSE}
plot(res)
```

Typical plots include:

- trajectories for different groups across `y0`, `y1`, `y2`
- observed vs expected post scores by baseline × trajectory group

These help you see at a glance which groups are doing better or worse than
expected.

# 5. Summaries and comparisons

The summary method provides a compact overview:

```{r}
summary(res)
```

You can also work directly with the core tables:

```{r}
res$Comparisons
res$GroupSummary
```

- `Comparisons` contains, for each baseline × trajectory group:
  - mean difference between observed and expected `y2`
  - t-statistics and p-values
- `GroupSummary` contains descriptive information (group sizes, means, etc.)

These tables are designed to be readable and easy to export to reports or
spreadsheets.

# 6. Conceptual sketch

SRTM is built around the idea that each student can serve as their **own
control**:

1. Use `y0` and `y1` to estimate each student's **personal trajectory** before
   the influence (their historical-to-baseline slope).
2. Assume that, over a short enough time window, this trajectory would have
   continued similarly if nothing new had happened.
3. Project this trajectory forward to obtain an **expected** `y2` for each
   student.
4. Compare the expected `y2` to the **observed** `y2`.
5. Group students by:
   - their pre-influence trajectory (e.g. declining, stable, improving), and
   - their baseline level (e.g. low, medium, high),
   then interpret and test differences at group level.

The goal is not to identify a single “best” practice, but to support teachers
and researchers in making **professional judgements** based on transparent,
locally meaningful evidence.

# 7. Next steps

From here you can:

- replace the example data with your own using `importSRTMExcel()`,
- change the grouping order via `group_first = "slope"`,
- explore plotting and reporting options tailored to your context.

For more detail on individual functions, see:

- `?SRTMAnalyse`
- `?compareOutcomes`
- `?summariseGroupOutcomes`
- `?SRTM_student_attitude_data` (example longitudinal dataset)

